Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-23 paper/paper.tex
some revision on Danvy's comments

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally to direct-style with delimited control. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the datatype representing continuations
  into higher-order functions.
  Following the idea that in deterministic languages, reduction contexts of reduction
  semantics are defunctionalized continuations, we further apply this idea to AAM --
  reduction contexts of it are refunctionalized to extended continuation-passing style.

@@

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination 
of abstractions to their concrete counterparts, abstract machines and definitional 
interpreters, respectively. An interesting question, and the subject of this paper, is 
how we can interderive these abstract semantics artifacts from the respective other one.

@@

In the domain of abstract semantics artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
The fundamental difference between concrete semantics artifacts and abstract semantics
artifacts is nondeterminism. In addition to ensuring termination, abstract semantics
artifacts are usually equipped with a cache of explored state that is guaranteed to reach
the least fixed-point eventually.
In this functional pearl, we provide a constructive answer to the question of interderiving
abstract semantics artifacts and relate AAM and ADI by presenting a series of syntactical
transformations on the program from small-step abstract abstract machines to big-step 
abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstracting} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [join={node[sloped,anchor=center,above,labeled] {\text{un-CPS}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearizing} \\ \text{+} \\ \text{fusing} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentangling}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantics artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

We begin by reviewing some background necessary for this work in Section~\ref{background}, 
as well as introducing some of the basic code structures used throughout the paper. 
We then address the main contribution of this paper, which is the filling of the gap between
small-step AAM and big-step ADI by developing a series of systematic transformations. 
Figure \ref{fig:trans} shows the transformations.

@@

  \item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation.
    The fusing transformation simply combines the single-step function @step@ and the 
    driving function into one, but keeps all the machine state representations.

  \item Section~\ref{disen} discusses \textbf{disentangling}, which disassembles the 
    fused AAM to be several individual functions, with each function handling one data 
    type represented by a continuation.

@@

% \begin{itemize}
% \item The main contribution of this paper is the filling of the gap
%   between small-step AAM and big-step ADI by developing a series of
%   systematic transformations.
%   We show that the correspondence between abstract
%   machines and interpreters exists not only in concrete semantics artifacts,
%   but also in abstract semantics artifacts.
%   Galois connection?
%   \note{JD: Section number?}

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, a control expression
to which the program may resume, and its environment.

@@

\item If the state is not a @Let@ expression, then it must be an atomic expression, as seen
  in the above code. In this scenario, we begin by extracting the top frame of all available
  continuations. The control (i.e., an atomic expression) of the current state is the 
  evaluated term that is being bound to the variable @x@ from the top frame.
  The environment and store are updated with @x@ mapping to the closure value of @ae@.
  Finally, the successor state is transferred to expression @e@ from the top frame,
  which is the body of a @Let@ expression, with the updated environment, store, and
  the rest of the stack @ks@.

@@

Similar to CESK machines, the machine state of AAM has a control expression,
an environment, a store, and continuation, as well as a timestamp.
However, there are several notable differences between AAM's store and CESK machine's store.
In AAM, the store maps addresses to sets of values; it stores all possible values
for a particular address. As such, dereferencing addresses becomes nondeterministic.
Also, the store performs \emph{joining}, rather than overwriting, when updating elements.
Furthermore, the continuations are likewise allocated on the store instead of formed into a
linked list, and the continuations are in a state which then turns into a continuation
address.

@@

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores,
nor handling the time stamp @Time@.
In abstract interpretation, however, these are key ingredients to achieve
analyses with different sensitivities, as well as to perform a finite
state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite
program contour @time@ that encodes the program execution history.
The function @tick@ is used to refresh the "time" and get a "new time".
We use a list of execution contexts (expressions) to represent the finite 
program contour, and as we will see in Section~\ref{kcfainst}, by applying 
different @tick@ functions on the timestamp, we are able to obtain a 
family of analyses.

@@

\begin{itemize}
  \item If the current control expression is a @Let@, then the result of @App(f, ae)@ will 
    be bound to variable @x@. In this case, we retrieve the set of closures that @f@ may 
    represent. For each closure in the set, we perform nearly the same operations as in the
    concrete CESK machines, with an important difference: the continuation is allocated on 
    the store @kstore@, so a new continuation address @new_kaddr@ must be constructed and 
    a new frame @Frame(x, e, env)@ paired with the current continuation address @kaddr@ is 
    merged into @new_kaddr@. Finally, a list of successor states is generated.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call stack as precise as runtimes
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match.
In this and following sections, we begin describing the transformations step by step.
Our base machine for now is the AAM with unbounded stack, though
as we will later transform the stack to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither requiring construction of a frame on stack,
nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

At this point, we have obtain a linearized abstract abstract machine. If we imagine that the
classical AAM explores a graph of reachable states, then the linearized AAM flattens the
graph to a linear sequence.

\section{Fusing} \label{fusing}

Fusing is a transformation that combines the @step@ and @drive@ functions into a single
function.
The fused function @drive_step@ is essentially formed by merging the functionality of
@step@ into the @drive@ function.
@drive_step@ takes an @NDState@ and a set of explored states as an argument
and returns a set of reachable states once it terminates.

@@

\section{Disentangling} \label{disen}

@@

The above code shows the skeleton of @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The astute reader will notice a vague shape of interpreter in continuation style has emerged.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely 
matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantics artifact, 
then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order nondeterministic 
evaluation through higher-order functions representing continuations, 
but also drives us to use a new caching algorithm.

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach
to approximate a program's runtime behavior\cite{cousot1977abstract}.
Control-flow analysis is one instance of abstract interpretation on functional programs
that can be traced to Jones \cite{jones1981flow}. \citeauthor{Shivers:1988:CFA:53990.54007} 
introduces $k$-CFA which uses $k$ recent calling contexts as program contour that 
differentiates values from different contexts \cite{Shivers:1988:CFA:53990.54007, 
Shivers:1991:SSC:115865.115884}.
The recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper.
Modern programming languages such as Java\cite{might2010resolving} and Racket 
\cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}
can also be modeled by the abstracting abstract machine approach. \\

\textit{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.
\citeauthor{Reynolds:72} first shows defunctionalization as a program transformation 
technique that can be used to transform higher-order functions to first-order functions
when specifying a programming language by an interpreter \cite{Reynolds:72, Reynolds:HOSC98-revisited}.
Refunctionalization is the left inverse of this transformation.

@@

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows
a reduction semantics is a structural operational semantics in continuation style,
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Our work further shows that for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifying and explicitly exposing 
(Section~\ref{linear}) that layer of nondeterministic continuations, the whole program can 
then be refunctionalized to extended continuation-passing style.

Fusion as one of the transformation in our paper is also studied by 
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}. \\

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.
Pushdown control-flow analysis (PDCFA) is a mechanism which maintains this precision
through the use of a Dyke state graph representing all possible stacks contained
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both
the source and target states of a call-site transition,
which guarantees that no spurious merging will occur during returns.

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines 
and big-step abstract definitional interpreters by developing a series of syntactical 
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally un-CPS with delimited control transforms
the abstract interpreter into a sequence of expressions, which looks no different to a
concrete interpreter.

We show the functional correspondence not only exists between concrete semantics artifacts
but also exists between abstract semantics artifacts. An interesting open question would
be whether there also exists a correspondence of static analyses formalized in different 
denotational style and operational style.

>>>>

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally to direct-style with delimited control. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the datatype representing continuations
  into higher-order functions.
  Following the idea that in deterministic languages, reduction contexts of reduction
  semantics are defunctionalized continuations, we further apply this idea to nondeterministic
  AAM -- reduction contexts of it are refunctionalized to extended continuation-passing style.

@@

Broadly speaking, abstract abstract machines and abstract definitional interpreters are
different forms of abstract interpreters. They are obtained by applying a combination 
of abstractions to their concrete counterparts, abstract machines and definitional 
interpreters, respectively. An interesting question, and the subject of this paper, is 
how we can interderive these abstract semantic artifacts from the respective other one.

@@

In the domain of abstract semantic artifacts, by contrast, the relations between small-step
abstract abstract machines and big-step abstract definitional interpreters, as well as the 
question of deriving one from the other, are not well studied.
One of the fundamental differences between concrete semantic artifacts and abstract semantic
artifacts is nondeterminism. In addition to ensuring termination, abstract semantic
artifacts are usually equipped with a cache of reachable states, the cache is
updated in a monotonic way and guaranteed to reach the least fixed-point eventually.
In this functional pearl, we provide a constructive answer to the question of interderiving
abstract semantic artifacts and relate AAM and ADI by presenting a series of syntactical
transformations on the program from small-step abstract abstract machines to big-step 
abstract definitional interpreters.

@@

\begin{figure}
\begin{center}
 \begin{tikzpicture}
   \matrix (m) [matrix of math nodes, row sep=2.5em, column sep=3.7em]
     {
         &  &  &  &
       \begin{smallmatrix} \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
       \begin{smallmatrix} \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
       \begin{smallmatrix} \text{disentangled}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{refunctionalized}  \\ \text{abstract} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{monadic} \\ \text{abstract} \\ \text{definitional} \\ \text{interpreters} \end{smallmatrix}  \\
       \begin{smallmatrix} \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{big-step} \\ \text{abstract} \\ \text{machines} \end{smallmatrix}  &
         &
       %\begin{smallmatrix} \text{disentangled} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
         &
       %\begin{smallmatrix} \text{refunctionalized} \\ \text{abstract} \\ \text{machines} \end{smallmatrix} &
       \begin{smallmatrix} \text{definitional} \\ \text{interpreters} \end{smallmatrix} \\
     };
   {
     [start chain] \chainin (m-2-4);
     \chainin (m-1-5) [join={node[sloped,anchor=center,above,labeled] {\text{un-CPS}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{}}}];
     \chainin (m-1-5) [join={node[above,labeled] {\text{}}}];
   }
   {
     [start chain] \chainin (m-2-1);
     \chainin (m-2-2) [join={node[above,labeled] {\begin{smallmatrix} \text{linearizing} \\ \text{+} \\ \text{fusing} \end{smallmatrix}}     }];
     \chainin (m-2-3) [join={node[above,labeled] {\text{disentangling}}}];
     \chainin (m-2-4) [join={node[above,labeled] {\text{refunc.}}}];
     \chainin (m-2-5) [join={node[above,labeled] {\text{monadify}}}];
   }
   {
     [start chain]
     \chainin (m-3-1);
     { [start branch=A] \chainin (m-2-1)
       [join={node[left, labeled] {\alpha}}];
     }
   }
     %\chainin (m-2-2);
     %\chainin (m-2-3);
     %\chainin (m-2-4);
   {
     [start chain]
     \chainin (m-3-5);
     { [start branch=B] \chainin (m-2-5)
       [join={node[right, labeled] {\alpha}}];
     }
     {
       [start branch=D, color=gray] \chainin (m-3-1)
       [join={node[right, labeled] {\text{}}}];
     }
   }
   {
     [start chain, color=gray] \chainin (m-3-1);
     {
       [start branch=C] \chainin (m-3-5)
       [join={node[above, labeled] {\text{Known as Danvy's correspondences for concrete semantic artifacts }}}];
     }
   }
 \end{tikzpicture}
 \caption{Transformations from AAM to ADI} \label{fig:trans}
\end{center}
\end{figure}

@@

We begin by reviewing some background necessary for this work in Section~\ref{background}, 
as well as introducing some of the basic code structures used throughout the paper. 
We then address the main contribution of this paper, which is the filling of the gap between
small-step AAM and big-step ADI by applying a series of well-known systematic transformations
drawn from concrete semantic artifacts.
Figure \ref{fig:trans} shows the transformations.

@@

  \item In Section~\ref{fusing}, we then present the \textbf{lightweight fusion} transformation.
    The fusing transformation simply combines the single-step function @step@ and the 
    driving function into one, but keeps all the machine state representations.

  \item Section~\ref{disen} discusses \textbf{disentanglement}, which disassembles the 
    fused AAM to be several individual functions, with each function handling one data 
    type represented by a continuation.

@@

% \begin{itemize}
% \item The main contribution of this paper is the filling of the gap
%   between small-step AAM and big-step ADI by developing a series of
%   systematic transformations.
%   We show that the correspondence between abstract
%   machines and interpreters exists not only in concrete semantic artifacts,
%   but also in abstract semantic artifacts.
%   Galois connection?
%   \note{JD: Section number?}

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, and a control component
to which the program may resume, and its environment.

@@

\item If the control component is not a @Let@ expression, then it must be an atomic expression,
  as seen in the above code. In this scenario, we begin by extracting the top frame of
  all available continuations. The variable @x@ from the top frame will be bound to
  the result of evaluating the atomic expression @ae@ by updating the environment and
  store.
  Finally, the successor state is transferred to expression @e@ from the top frame,
  which is the body of a @Let@ expression, with the updated environment, store, and
  the rest of the stack @ks@.

@@

Similar to CESK machines, the machine state of AAM has a control, an environment, a store,
and continuation, as well as a timestamp. However, there are several notable differences
between AAM's store and CESK machine's store. In AAM, the store maps addresses to sets of
values; it stores all possible values for a particular address. As such, dereferencing
addresses becomes nondeterministic. Also, the store performs \emph{joining}, rather than
overwriting, when updating elements. Furthermore, the continuations are likewise allocated
on the store instead of formed into a linked list, and the continuation component becomes
an address that maps to a set of continuations in the store instead of directly embedded
in the state.

@@

\subsubsection{Allocating Addresses}
Up to this point, we have not described allocating addresses in stores,
nor handling the time stamp @Time@.
In abstract interpretation, however, these are key ingredients to achieve
analyses with different sensitivities, as well as to perform a finite
state space analysis\cite{Gilray:2016:ACP:2951913.2951936}.
To effectively approximate the runtime behavior, we introduce a finite
program contours @time@ that encodes the program execution history.
The function @tick@ is used to refresh the "time" and get a "new time".
We use a list of expression (which are drawn from the control component)
to encode the calling contexts history, and as we will see in Section~
\ref{kcfainst}, by applying different @tick@ functions on the timestamp,
we are able to obtain a family of analyses.

@@

\begin{itemize}
  \item If the current control component is a @Let@, then the result of @App(f, ae)@ will 
    be bound to variable @x@. In this case, we retrieve the set of closures that @f@ may 
    represent. For each closure in the set, we perform nearly the same operations as in the
    concrete CESK machines, with an important difference: the continuation is allocated on 
    the store @kstore@, so a new continuation address @new_kaddr@ must be constructed and 
    a new frame @Frame(x, e, env)@ paired with the current continuation address @kaddr@ is 
    merged into @new_kaddr@. Finally, a list of successor states is generated.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding addresses of 
continuations in states, we intend to use a list of frames to explicitly model 
the stack. By doing so, we recover the call stack as the same as runtimes
(so called pushdown control-flow analysis), but since the stack is unbounded, 
the state space is potentially infinite.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match. Now we begin describing the
transformations step by step. Our base machine for now is the AAM with unbounded
stack, though as we will later transform the stack to higher-order functions
representing continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither requiring construction of a frame on stack,
nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because
it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

At this point, we have obtained a linearized abstract abstract machine. If we imagine that the
classical AAM explores a graph of reachable states, then the linearized AAM flattens the
graph to a linear sequence.

\section{Lightweight Fusion} \label{fusing}

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ is essentially formed by merging the functionality of
@step@ into the @drive@ function.
@drive_step@ takes an @NDState@ and a set of explored states as an argument
and returns a set of reachable states once it terminates.

@@

\section{Disentanglement} \label{disen}

@@

The above code shows the skeleton of @drive_step@.
At the end of the first case of pattern matching, we invoke @drive_step@ recursively 
\footnote{The code for constructing new environments, stores, and frames is elided for 
simplicity of presentation.}.
The code for the second case is replaced entirely by a function call to @continue@.
The shape of an interpreter in continuation-passing style has emerged.
%The astute reader will notice a vague shape of interpreter in continuation style has emerged.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely 
matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantic artifact, 
then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order nondeterministic 
evaluation through higher-order functions representing continuations, 
but also drives us to use a new caching algorithm.

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach
to approximate a program's runtime behavior\cite{cousot1977abstract}.
Control-flow analysis is one instance of abstract interpretation on functional programs
that can be traced to Jones \cite{jones1981flow}. \citeauthor{Shivers:1988:CFA:53990.54007} 
introduced $k$-CFA which uses $k$ recent calling contexts as program contours that 
differentiate values from different contexts \cite{Shivers:1988:CFA:53990.54007, 
Shivers:1991:SSC:115865.115884}.
The recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper.
Modern programming languages such as Java\cite{might2010resolving} and Racket 
\cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}
can also be modeled by the abstracting abstract machine approach. \\

\textit{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.
\citeauthor{Reynolds:72} first shows defunctionalization as a program transformation 
technique that can be used to transform higher-order functions to first-order functions
when writing a definitional interpreter \cite{Reynolds:72, Reynolds:HOSC98-revisited}.
Refunctionalization is a left inverse of this transformation.

@@

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows
a reduction semantics is a structural operational semantics in continuation style,
where the reduction context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
% Danvy also shows in nondeterministic languages ...
Our work further shows that for nondeterministic languages, there is another layer of continuations
that controls the nondeterministic choices. After identifying and explicitly exposing 
(Section~\ref{linear}) that layer of nondeterministic continuations, the whole program can 
then be refunctionalized to extended continuation-passing style.

Lightweight fusion as one of the transformation in our paper is also studied by 
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}. \\

@@

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to having an exponential time
complexity.
Pushdown control-flow analysis (PDCFA) is a mechanism which maintains this precision
through the use of a Dyke state graph representing all possible stacks contained
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both
the source and target states of a call-site transition,
which guarantees that no spurious merging will occur during returns.

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines 
and big-step abstract definitional interpreters by applying a series of syntactical 
transformations. Among these transformations, linearization turns a worklist into another
layer of continuations; refunctionalization converts the first-order data types representing
continuations to higher-order functions; and finally un-CPS with delimited control transforms
the abstract interpreter into a sequence of expressions, which looks no different to a
concrete interpreter.

We show the functional correspondence not only exists between concrete semantic artifacts
but also exists between abstract semantic artifacts. An interesting open question would
be whether there also exists a correspondence of static analyses formalized in different 
denotational style and operational style.
